---
title: "Discussion1"
output: html_document
date: "2023-04-06"
editor_options: 
  markdown: 
    wrap: 72
---

# Discussion 1

#### April 7, 2023

TA email: [nylli\@ucdavis.edu](mailto:nylli@ucdavis.edu){.email}

## Overview

1.  OMSI system testing
2.  New feature introduction: chat
3.  R for basic linear algebra
4.  'Compas' dataset
5.  'German credit' dataset

## 1. OMSI

### Setup:

-   Download OMSI
-   Connect to library VPN
-   Go to OMSI dir, find "OmsiGui.py"
-   Run "OmsiGui.py" on terminal with cmd line script:

### Run OMSI:

-   **Command format:** python3 OmsiGui.py (localhost) (port#) (your ucd
    email) (quiz name)

-   **Example:** python3 OmsiGui.py pc11.cs.ucdavis.edu 4000
    [nylli\@ucdavis.edu](mailto:nylli@ucdavis.edu){.email} quiz0_sq23

-   **On my machine:** cd "C:/Program Files/omsi-master" python
    OmsiGui.py pc11.cs.ucdavis.edu 4000
    [sydwoods\@ucdavis.edu](mailto:sydwoods@ucdavis.edu){.email}
    quiz0_sq23

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 2. Linear Algebra

```{r}
# Matrix
A <- matrix(c(2,2,3,4), nrow=2, ncol=2)

# Inverse 
inv(A)

# Eigenvalues and Eigenvectors
eigen(A)
eig$values
eig$vectors

# Row bind
rbind(A,B)

```

## 3. COMPAS Data Set

```{r}

```

## 4. German Credit Data Set

```{r}

```
