---
title: "HW2"
output: html_document
date: "2023-05-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problem 1

SAT Variables

-   **SAT score (Proxy)** - use percentile and then cut to separate into nQuantl = 5 quantiles

-   **Family income (Cat)** - use cut to separate into nCut = 4 groups (unless categorical)

-   **College acceptance (Y)** - predicting with modification of SAT score based on family income group

PEF Variables

-   **Weeks worked (Proxy)** - use percentile and then cut to separate into nQuantl = 5 quantiles

-   **Age (Cat)** - use cut to separate into nCut = 4 groups (unless categorical)

-   **Wage income (Y)** - predicting with modification of wkswrkd based on age group

## Cut Function

```{r}
library(qeML)
data(pef)

Ncut <- function(data, Name, numCuts=2){
  if (!is.numeric(data[,Name])){
    stop("data must be numeric, call form: cutN(data, categoryName, numCuts)")
  }
  df <- data[order(data[,Name]),,drop=FALSE]  # order by wkswrkd, keep original row indices to sort later
  size <- length(df[,Name])                   # numrows = 20090
  intervalSize <- ceiling(df[size, Name]/ numCuts) # interval size 13 for splitting min:max 0:52 wkswrkd into 4 categories
  iVal <- intervalSize
  cutPoints <- c(1)
  for (i in 1:numCuts){                       # 1:4 groups of wkswrkd
    val <- match(iVal, df[,Name])             # find group cutoff (i.e. 13) index
    if (is.na(val)){                          # if no match, find closest neighbor index (min)
      val <- which.min(abs(df[,Name] - iVal)) 
    }
    if (i == numCuts){                        # for last group: add last row index to index list, exit
      cutPoints <- append(cutPoints, size)
      break
    }
    cutPoints <- append(cutPoints, val)       # add last row index to index list
    iVal <- iVal + intervalSize               # i.e. 13+13 -> 26+13 -> 39+13 -> 52
  }
  print(cutPoints)                            # print indices of cutoffs
  df$newFactors <- ""
  for (i in 1:numCuts) {
    first <- df[cutPoints[i], Name]           # min of range (i.e. 13 wkswrkd)
    second <-df[cutPoints[i+1], Name]         # max of range (i.e. 26 wkswrkd)
    label <- paste("(", first, ", ", second, "]", sep = "") # i.e. "(13, 26]"
    indexMin <- cutPoints[i]                  # index of min (i.e. 1643)
    indexMax <- cutPoints[i+1]                # index of max (i.e. 2310)
    df$newFactors[indexMin:indexMax] <- label # add range label to new column
  }
  df$newFactors <- as.factor(df$newFactors)   # convert to factor
  print(unique(df$newFactors))
  df <- df[order(as.numeric(row.names(df))),] # sort using original order
  return(df$newFactors)
}

data$wkswrkdCategory <- Ncut(pef, 'wkswrkd', 4) # test!
head(data)
```

## 1. Generate Percentiles for Proxy (wkswrkd, SAT score)

```{r}
data(pef)
data <- pef         # data
proxName <- 'wkswrkd'   # X value (i.e. SAT score)

data$proxPercentile <- rank(data[,proxName])/nrow(data)*100
head(data)
```

## 2. Cut and Label Data by Proxy Quantile

```{r}
nQuantls <- 2
data$proxCategory <- Ncut(data,"proxPercentile", numCuts=nQuantls)
head(data)
```
