---
title: "L3"
output: html_document
date: "2023-04-11"
---

### April 6, 2023

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Simpson's Paradox

Two vars have positive relation:

-   (+) rel(x, y)

But the same vars have a negative relation when conditioned on 3rd var:

-   (-) rel(x, y \| z)

### Load UCB Data

```{r}
data("UCBAdmissions")
dim(UCBAdmissions)
dimnames(UCBAdmissions)
```

### Determine object structure:

**Structure:** This data is a 3D table with various levels per attribute

```{r}
str(UCBAdmissions)
```

**Margin table:** temporarily ignoring department attribute, find each variable's "marginal counts" irrespective of department

```{r}
UCB.GA <- margin.table(UCBAdmissions, c(1,2))
UCB.GA
```

```{r}
univ <- apply(UCBAdmissions, c(1,2), sum)
univ
```

Note: on the surface women appear to be proportionally much more frequently rejected than men

```{r}
prop.table(univ, 2)
```

Tool to develop: input data set, output graphically whether paradox applies

Consider the number of unique values of x or y; if 2 then variable is dichotomous. For dichotomous variables, perform alternative (Kendall's) correlation measure; for non-dichotomous variables, use classical correlation.

Graphical output:

-   Built-in base R graphics: plot(u,v,col=w) \# col is hue, barplot

-   Ggplot2 (ECS 132 appendix resource)

-   Lattice

Extra: find a data set & variables where paradox applies This data set is discrete, but can be applied to continuous data

```{r}
library(vcd)
```

## Fourfold Chart

Invented by Edward Tufte

Shape:

-   Top row: admitted (L) and rejected (R) men

-   Bottom row: admitted (L) and rejected (R) women

Features:

-   In any 2 adjacent quadrants, the ratio of the areas shown is the oz ratio (Opposite of pi chart, constant angle with varying radius, whereas pi chart is varying angle with constant radius)

-   95% confidence interval around edge

    <div>

    -   $S.E. (\overline x ) = S / \sqrt(n)$

    -   $\overline x +/- 1.96*(S.E.(\overline x))$

    </div>

### Compare total acceptance

```{r}
UCB <- aperm(UCBAdmissions, c(2,1,3))
fourfold(margin.table(UCB, c(1,2)))
```

### Compare chart by department

Aside from top-left chart, departments have nearly identical quadrants (similar M/F admission per dept)

Top-left chart: admissions favored women (admitted women at higher rates than men)

```{r}
fourfold(UCBAdmissions, mfrow=c(2,3))
```

### Compare admission rates by department

```{r}
ftable(round(prop.table(UCBAdmissions, c(2,3)),2),
       row.vars="Dept", col.vars=c("Gender","Admit"))
```

Key terms to search:

-   Contingency tables

-   Log-linear model

-   X - explanatory variable(s), predictor(s), feature(s)

-   Y - response variable, outcome variable, target

Example:

-   X: weeks worked

-   Y: wage income

-   Z: age

-   Vertical axis will be correlations (find for z=1, z=2, ...)

-   Graph correlations: find correlations for each z, then graph

-   Bar heights are counts of correlations

-   Number of bins for z (age)?

```{r}
UCB_G <- structable(Gender - Dept + Admit, data=UCBAdmissions)
cotabplot(UCB_G)
```
