---
title: "L4"
output: html_document
date: "2023-04-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(qeML)
data(mlb)
head(mlb)
```

## Notation

**Y**: what we are predicting

**X**: what we are using to make predictions

**Indicator/Dichotomous Variables:** (In R) 2-level factors typically 1/0 Coded

-   X features can be indicators (dummy variables)

-   Y can be a single indicator or a set of indicator variables

**Arguments:**

-   Formal arg: \<y name\>

<!-- -->

-   Actual arg: 'Weight'

-   Optional args: have default values or NA

### Example: KNN, Random Forest Train

-   Call form: `model fit <- qe<model name>(<data name>,<Y name>)`

-   Call:

```{r}
mlb1 <- mlb[,4:6]  # columns for height, weight and age
knnout <- qeKNN(mlb1,'Weight')  # fit k-Nearest Neighbor model
rfout <- qeRF(mlb1,'Weight')  # fit random forests model
```

### Example: Random Forest Predict

-   Call form: `predict(<model fit>, <new X value>)`

-   Call:

```{r}
predict(rfout,c(70,28))
```

### KNN Error (MAE)

```{r}
knnout$testAcc
```

### Random Forest Error (MAE)

```{r}
rfout$testAcc
```

**Default args in R:** Ideal for versatile user base

**Hyper parameters:** i.e. tuning parameters

-   Ex. number of nearest neighbors (e.g. k = 5 will use closest 5 instances to predict new instance)

### Regression and Classification

**Regression Settings:** y is continuous or ordinal

**Classification Settings:** y is categorical

-   Y is converted to a set of indicator variables

-   $E(Y | X = t)$ reduces to $P(Y = 1 | X = t)$

-   Find probabilities for each category, then guess the one with the highest probability.

**The Regression Function:** $m(t) = E(Y | X = t)$

-   $m(t)$: the mean value of $Y$ in the subpopulation defined by $X = t$

-   Ex. $m(71,23)$ would be the mean weight among all players of height 71 inches and 23 years old

-   Both regression and classification tasks make use of the regression function
