---
title: "HW1Problem1"
output: html_document
date: "2023-05-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(qeML)
data(pef)
library(fairml)
data(law.school.admissions)
```

### Cut Function

Inputs:

-   **data**

-   **Name** - variable to partition data by

-   **numCuts** - number of partitions to make

Outputs:

-   **df\$newFactors** - column with categorical labels (i.e. (13, 26])

```{r}
# To be applied to percentiles, to cut neatly into 0-20, 20-40, 40-60, etc. bins
Ncut <- function(data, Name, numCuts=2, max=NULL, min=NULL){
  if (!is.numeric(data[,Name])){
    stop("data must be numeric, call form: cutN(data, categoryName, numCuts)")
  }
  df <- data[order(data[,Name]),,drop=FALSE]  # order by wkswrkd, keep original row indices to sort later
  size <- length(df[,Name])                   # numrows = 20090
  
  if (is.null(max)) {
    intervalSize <- ceiling(df[size, Name]/ numCuts) # ival size 13 for splitting min:max 0:52 wkswrkd into 4 categories
  } else {
    intervalSize <- (max-min)/numCuts # ival size 25 for splitting percentiles into 4 groups
  }
  
  iVal <- intervalSize
  cutPoints <- c(1)
  for (i in 1:numCuts){                       # 1:4 groups of income percentile
    val <- match(iVal, df[,Name])             # find group cutoff (i.e. 20th) index
    if (is.na(val)){                          # if no match, find closest neighbor index (min)
      val <- which.min(abs(df[,Name] - iVal)) 
    }
    if (i == numCuts){                        # for last group: add last row index to index list, exit
      cutPoints <- append(cutPoints, size)
      break
    }
    cutPoints <- append(cutPoints, val)       # add last row index to index list
    iVal <- iVal + intervalSize               # i.e. 0+25 -> 25+25 -> 50+25 -> 75+25 -> 100
  }
  print(cutPoints)                            # print indices of cutoffs
  df$newFactors <- ""
  for (i in 1:numCuts) {
    if (is.null(max)) {
      first <- df[cutPoints[i], Name]         # min of range (i.e. 13 wkswrkd)
      second <-df[cutPoints[i+1], Name]       # max of range (i.e. 26 wkswrkd)
    } else {
      first <-  min+(intervalSize*(i-1))      # min of PERCENTILE range (i.e. 0)
      second <- min+(intervalSize*(i))        # max of PERCENTILE range (i.e. 20)  
    }
    
    label <- paste("(", first, ", ", second, "]", sep = "") # i.e. "(13, 26]"
    indexMin <- cutPoints[i]                  # index of min (i.e. 1643)
    indexMax <- cutPoints[i+1]                # index of max (i.e. 2310)
    df$newFactors[indexMin:indexMax] <- label # add range label to new column
  }
  df$newFactors <- as.factor(df$newFactors)   # convert to factor
  print(unique(df$newFactors))
  df <- df[order(as.numeric(row.names(df))),] # sort using original order
  return(df$newFactors)
}
```

### Cut Function Example

```{r}
data <- pef
Name <- 'wkswrkd'
numCuts <- 4
data$wkswrkdCategory <- Ncut(data=data, Name=Name, numCuts=numCuts)
head(data[,c('wkswrkd','wkswrkdCategory')], 10)
```

# 1. Relative Proxy Function

Inputs:

-   **data** - a data frame or equivalent

-   **proxyName** - e.g. SAT score in the above example (proxy for wealth)

-   **catName** - e.g. family income in the above example; presumed to be a categorical variable, i.e. an R factor

-   **nQuantls** - the number of percentiles we'll use

-   **nCut** - applies to the case in which catName is originally a continuous variable; it will be tranformed to an R factor by converting it to this many intervals, as in the R cut() function

Outputs:

-   The return value will be an object of class 'relProx', an R list quantls with one component for each level in the catName column (after cut()-like operatin, if any). Said component will consist of the nQuantls of the proxyName column within the rows of data for that category.

-   The return value will also need to record the intervals made by the cut()-like function, if any (see below). Do this by attaching an R 'attribute' to the return value.

### RelProx Implementation

```{r}
relativeProxy <- function(data,proxyName,catName,nQuantls,nCut=NULL) {
  if (is.null(nCut)) {
    # Split 'income' by categorical levels
    data$catCategory <- data[,catName]
  } else {
    # Split 'income' into groups by percentile if continuous (otherwise, leave as is)
    data$catPercentile <- rank(data[,catName])/nrow(data)*100
    data$catCategory <- Ncut(data,"catPercentile", numCuts=nCut)
  }
  
  # Create S3 list of dataframes split by category levels
  catLevels <- levels(data$catCategory)
  quantls <- split(data,data$catCategory,drop = FALSE)
  class(quantls) <- 'relProx'
  
  # In each category, find relative percentiles and quantiles for proxy variable
  for (i in 1:length(quantls)) {
    # Isolate dataframe for a particular level
    subset <- quantls[[i]]
    
    # Calculate percentiles for each proxy value
    subset$proxyPercentile <- rank(subset[,proxyName])/nrow(subset)*100
    
    # Split proxy into quantiles by percentile
    subset$proxyCategory <- Ncut(subset,"proxyPercentile", numCuts=nQuantls, max=100, min=0)
    
    # Re-assign dataframe with new columns
    quantls[[i]] <- subset
  }
  return(quantls)
}
```

### RelProx Example: Continuous Category

```{r}
data <- pef              # data
proxyName <- 'wkswrkd'   # proxy (i.e. wkswrkd, SAT score)
catName <- 'age'         # category (i.e. age, income)
nQuantls <- 5            # split 'wkswrkd' into two high/low groups
nCut <- 4                # split 'age' into four groups

quantls <- relativeProxy(data=data,proxyName=proxyName,catName=catName,nQuantls=nQuantls,nCut=nCut)
head(quantls[[1]])
```

### RelProx Example: Categorical Category

```{r}
data <- law.school.admissions[1:20790,]
proxyName <- 'lsat'      # proxy
catName <- 'fam_inc'     # category
nQuantls <- 5            # split 'lsat' into 4 groups
quantls <- relativeProxy(data=data,proxyName=proxyName,catName=catName,nQuantls=nQuantls)
head(quantls[[1]])
```

## 2. Predict Function

`predict <- function(relProxObj, newCases, proxyName, catName) {}`

### Merge data after getting relProxObj from relativeProxy()

```{r}
mergedData <- quantls[[1]]    # merge all split data together
for (i in 2:length(quantls)) {
  mergedData <- rbind(mergedData,quantls[[i]])
}

head(mergedData)
```

### Train qeKNN model on merged data

> Use only the variables **fam_inc**, **lsat**, **ugpa**, **cluster** and **bar**, with that last one being the outcome ("Y") variable.

```{r}
yName <- 'bar'
adjustedData <- mergedData
adjustedData$lsat <- adjustedData$proxyCategory
adjustedData <- adjustedData[,c('fam_inc','lsat','ugpa','cluster','bar')]

model <- qeKNN(data=adjustedData, yName=yName)  # train qeKNN on ALL data
```

### Determine relative quantile of new data

-   Determine the income group of the new person

-   Find LSAT score percentile relative to income group

-   Use trained knn model to predict "bar" with modified LSAT score group

```{r}
relProxObj <- quantls
newCases <- law.school.admissions[20791:20800,!names(data) %in% c(yName)]
newCases <- newCases[,c('fam_inc','lsat','ugpa','cluster')]
rownames(newCases) <- 1:nrow(newCases) # drop target and re-index for new cases

proxyName <- 'lsat'
catName <- 'fam_inc'
colnames(newCases)
```

### Predict BAR with trained qeKNN using adjusted score

```{r}
newCaseCats <- c()
for (i in 1:nrow(newCases)) {
  case <- newCases[i,]
  caseCat <- case[,catName]
  caseProxy <- case[,proxyName]
  
  # TODO: COMPARE TO THRESHOLD STORED SOMEWHERE
  # Determine which df in relProxObj to compare to
  df <- relProxObj[[1]] #########################
  
  
  proxyList <- c(df[,proxyName], caseProxy)
  proxyListLen <- length(proxyList)
  percentile <- (rank(proxyList)/proxyListLen*100)[proxyListLen]
  
  # Determine which quantile the score falls within
  numQuants <- length(levels(df$proxyCategory)) # number quantiles
  iVal <- 100 / numQuants # size of range
  mult <- caseProxy %/% iVal # use mod to find category
  pQuantMin <- iVal*(mult)
  pQuantMax <- iVal*(mult+1)
  label <- paste("(",pQuantMin,", ",pQuantMax,"]",sep="") # i.e. "(20, 40]"
  percentileQuant <- as.factor(label)
  case$lsat <- percentileQuant
  caseCat <- predict(model, case)$predClasses[1,1]
  newCaseCats <- c(caseCat,newCaseCats) # predict class of new cases
  
  return(newCaseCats)
}
```
