---
title: "Quiz0"
output: html_document
date: "2023-04-10"
---
```{r}
.libPaths(c("C:/Users/Sydney/Documents/R/win-library/4.1", "C:/Program Files/R/R-4.1.3/library"))
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1:

Consider the mtcars dataset, built-in to R. Write code that finds and prints the hp value in the third row of the dataset.

```{r}
# Select 3rd value of hp column and print
print(mtcars$hp[3])
```

## Question 2:

Again with the mtcars dataset, write code that finds and prints the mean mpg among cars having 6 or 8 cyclinders.

```{r}
# Create subset of mtcars with 6 or 8 cylinders
subset <- mtcars[which(mtcars$cyl == 6 | mtcars$cyl == 8),]
# Print mean mpg of this subset
print(mean(subset$mpg))
```

## Question 3:

Again with the mtcars dataset, write code that adds a new column, the horsepower/weight ratio, and prints the entire new dataset, now 12 columns rather than 11.

```{r}
# Create ratio column by dividing hp/wt
mtcars$ratio <- mtcars$hp/mtcars$wt
# Print full dataset
print(head(mtcars))
```

## Question 4:

Write code that prints: the proportion of foreign workers who have good credit; the proportion of workers with good credit who are foreign; the proportion of foreign workers; and the proportion of workers who have good credit. (Print a 4-element vector.)

```{r}
# Load fairml for german.credit dataset
library(fairml)

# Find credit of all workers, foreign workers, non-foreign workers
credit_all <- german.credit$Credit_risk # n = 1000
credit_foreign <- german.credit$Credit_risk[which(german.credit$Foreign_worker == "yes")] # n = 963
credit_nonforeign <- german.credit$Credit_risk[which(german.credit$Foreign_worker == "no")] # n = 37

# P(foreign workers who have good credit) = # foreign with good credit / # foreign
p_foreign_good <- length(credit_foreign[which(credit_foreign == "GOOD")]) / length(credit_foreign)

# P(workers with good credit who are foreign) = # foreign workers with good credit / # workers with good credit
p_good_foreign <- length(credit_foreign[which(credit_foreign == "GOOD")]) / length(credit_all[which(credit_all == "GOOD")])

# P(foreign workers) = # foreign workers / # workers
p_foreign <- length(credit_foreign) / length(credit_all)

# P(workers who have good credit) = # good credit workers / # workers
p_good <- length(credit_all[which(credit_all == "GOOD")]) / length(credit_all)

# Combine answers into vector and print
answers <- c(p_foreign_good, p_good_foreign, p_foreign, p_good)
print(answers)
```

## Subset Data Frames

```{r}
head(german.credit[,c("Credit_risk","Foreign_worker")])
```

```{r}
head(german.credit[which(german.credit$Foreign_worker == "yes"),c("Credit_risk","Foreign_worker")])
```


```{r}
head(german.credit[c("Credit_risk","Foreign_worker")])
```


```{r}
head(german.credit[which(german.credit$Foreign_worker == "yes"),])
```
