---
title: "HW3md"
output: html_document
date: "2023-05-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(qeML)
library(fairml)
library(dplyr)
library(ggplot2)
data(compas)
compas$two_year_recid <- as.numeric(compas$two_year_recid == 'Yes')

data <- compas
yName <- 'two_year_recid'
sName <- 'race'
xName <- 'age'
condits <- c('priors_count <= 4','decile_score >= 6')
qeFtn <- qeKNN
minS <- 50
yLim <- NULL
useLoess <- TRUE
```

Apply conditions to data

```{r}
data <- compas
for (i in 1:length(condits)) { 
  condit <- paste("data[which(data$",condits[i],"),]",sep="")
  expr <- parse(text=condit)
  data <- eval(expr)
} # IT WORKS
```

Train qeFtn on data

```{r}
model <- qeFtn(data,yName)
```

Create categories for x (i.e. age): either unique values, or quantiles/some other method if continuous

-   Option A: unique values (use this for age)

-   Option B: make quantile cutoffs (helper function getDiv chooses an ideal number of quantiles so that on average, there will be 10 people per category

```{r}
# OPTION B
# Find ideal even number of x quantiles with enough ppl per s-group
getDiv <- function(data, xName, dataLen) {
  div <- 0
  ratio <- 0
  while(ratio < 20) { # We want at least a 10:1 ratio of data rows:categories
    div <- div + 1
    newCol <- data[[xName]] %/% div
    numCats <- length(unique(newCol))
    ratio <- dataLen/numCats
  }
  return(div)
}
```

```{r}
# OPTION B
# Find percentile-based quantiles for x categories; return column
getQuants <- function(data, xName, dataLen) {
  quantSize <- getDiv(data, xName, dataLen)
  quantCol <- ntile(data[,xName], quantSize)
  return(quantCol)
}
```

Get recidivism/y probabilities for each age category

TODO: collect and plot

```{r}
dataLen <- length(data[[xName]])                # number rows
uniqueXNum <- length(unique(data[[xName]]))     # number unique x values (i.e. ages)

useUnique <- TRUE
#useUnique <- ifelse (dataLen/uniqueXNum > 10, TRUE, FALSE) # use unique values if each category has, on average, 10 data points

# Define category column
if (useUnique) { # Option A
  data$xCats <- data[[xName]]
} else {         # Option B
  data$xCats <- getQuants(data, xName, dataLen)
  #xRng <- max(data[[xName]]) - min(data[[xName]]) # i.e. 49
  #data$xCats <- cut(data[[xName]], quantile(data[[xName]],probs=seq(0,1,1/quantSize),na.rm=TRUE), include.lowest=TRUE,labels=FALSE)
}

# Get unique categories as list
xCats <- unique(data$xCats)
sCats <- unique(data[,sName])
sCatPreds <- data.frame(matrix(nrow=length(xCats),ncol=length(sCats)+1))
sCatPreds[,1] <- xCats # Column 1: x category

# Subset data based on category, find mean probability for cat based on model preds
for (i in 1:length(xCats)) {
  x <- xCats[i]
  xSubset <- data[which(data$xCats==x),]
  xSubset <- xSubset[,!(names(xSubset) %in% c(yName,'xCats'))]
  
  #allPreds <- predict(model, xSubset)
  #avgAllPreds <- mean(allPreds)
  
  predList <- c(x)
  for (j in 1:length(sCats)) { # fill in remaining columns with s-respective mean preds
    sCat <- sCats[j]
    xsSubset <- xSubset[which(xSubset[ ,sName]==sCats[j]),]
    if (nrow(xsSubset) > 0) {
      sPreds <- predict(model, xsSubset)
      sPredsList <- sPreds[1,]
      avgSPreds <- mean(sPredsList)
      #predList <- c(predList,avgSPreds)
      sCatPreds[i,(j+1)] <- avgSPreds
    }
  }
}

# TODO: convert numeric encoding to race names for the plot
colnames(sCatPreds) <- c('xCat', sCats)
```

Plot

TODO: NEEDS SMOOTHING, DESPARATELY

TODO: add race labels based

```{r}
predsMelt <- melt(sCatPreds,id.vars='xCat', variable.name = 'series')

#create line plot for each column in data frame
ggplot(predsMelt, aes(xCat, value)) +
  geom_line(aes(colour = series))
```

Alternative Plotting (base R)

```{r}
plot(sCatPreds[c(1,2)], col="red")

points(sCatPreds[c(1,3)], col="orange")
points(sCatPreds[c(1,4)], col="yellow")
points(sCatPreds[c(1,5)], col="green")
points(sCatPreds[c(1,6)], col="blue")
points(sCatPreds[c(1,7)], col="purple")
```

Loess Smoothing

```{r}
LoessDf <- sCatPreds
LoessDf <- LoessDf[order(LoessDf$xCat),]

colorList <- rainbow(length(colnames(LoessDf))-1)

#xLim <- c(min(LoessDf$xCat), max(LoessDf$xCat))
#yLim <- c(0, 100)

plot(LoessDf$xCat, LoessDf$y, pch=19, main='Race-Stratified Recidivism Predictions for each Age Group', col="white")
#plot.window(xlim=xLim, ylim=yLim)
for (i in 2:(length(colnames(LoessDf)))) {
  yColName <- colnames(LoessDf)[i]
  TestDf <- LoessDf[,c('xCat',yColName)]
  TestDf$y <- LoessDf[,yColName]
  TestDf <- na.omit(TestDf)
  
  loess50 <- loess(xCat ~ y, TestDf, span=0.8)
  smooth50 <- predict(loess50)
  lines(smooth50, x=TestDf$xCat, col=colorList[i-1])
}

legend('bottomright', legend=sCats,
        col=colorList, pch=19, title='Sensitive Variable Groups')
```
